---
title: "*Ex situ* conservation gap analysis summary"
subtitle: "`r params$taxon`"
author: "Generated with GAMMa (Gap Analysis &amp; Metacollection Management) v`r params$app_version`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    self_contained: true
params:
  taxon: "Taxon"
  app_version: "dev"
  points: NA
  records: NA
  bufferDist: NA
  method: "buffer"
  sf_model: NA
  srsMetrics: NA
  grsMetrics: NA
  ersMetrics: NA
  fcs: NA
  priority: "Not assessed"
  priorityColor: "#dddddd"
  sf_buffers: NA
  sf_grs_gap: NA
  sf_ers_regions: NA
  combinedColor: NA
  grsexColor: NA
  ersexColors: NA
---

```{=html}
<style>
  /* System fonts: the report must read the same offline */
  body { font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, "Helvetica Neue", Arial, sans-serif; }
  div.main-container { max-width: 1400px !important; }
  h1.title { font-size: 1.9rem; margin-bottom: 0.2rem; }
  h3.subtitle { font-size: 1.5rem; font-style: italic; color: #2c3e50; margin-top: 0; }
  h4.author, h4.date { font-size: 0.95rem; font-weight: 400; color: #6c757d; margin: 0.15rem 0; }
  h1 { font-size: 1.5rem; margin-top: 2.5rem; padding-bottom: 0.4rem; border-bottom: 2px solid #18bc9c; }
  h2 { font-size: 1.2rem; margin-top: 1.8rem; color: #2c3e50; }
  h3 { font-size: 1.05rem; margin-top: 1.4rem; }
  p, li { font-size: 0.95rem; line-height: 1.55; }
  .score-cards { display: flex; flex-wrap: wrap; gap: 14px; margin: 1rem 0 1.5rem; }
  .score-card { flex: 1 1 150px; border-radius: 8px; padding: 14px 16px; color: #fff; box-shadow: 0 2px 6px rgba(0,0,0,.12); }
  .score-card .label { font-size: 0.8rem; text-transform: uppercase; letter-spacing: .05em; opacity: .9; }
  .score-card .value { font-size: 2rem; font-weight: 700; line-height: 1.1; }
  .score-card .sub { font-size: 0.8rem; opacity: .9; }
  .priority-card { color: #2c3e50; border: 2px solid rgba(0,0,0,.08); }
  .facts { display: grid; grid-template-columns: repeat(auto-fit, minmax(210px, 1fr)); gap: 8px 24px; margin: .5rem 0 1rem; font-size: .92rem; }
  .facts div span { color: #6c757d; display: block; font-size: .78rem; text-transform: uppercase; letter-spacing: .04em; }
  .note { background: #f6f8fa; border-left: 4px solid #18bc9c; padding: .6rem .9rem; margin: .8rem 0; font-size: .9rem; }
  .leaflet-popup-content { max-width: 280px; word-wrap: break-word; overflow-wrap: anywhere; }
  table.table { font-size: .9rem; }
  .footer-note { font-size: .8rem; color: #6c757d; border-top: 1px solid #dee2e6; margin-top: 2.5rem; padding-top: .8rem; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Explicit library() calls so rsconnect / renv detect the dependencies
library(leaflet)
library(dplyr)
library(tidyr)
library(DT)
library(sf)
library(htmltools)
library(rmapshaper)
library(plotly)
library(kableExtra)

combinedColor <- params$combinedColor
grsexColor    <- params$grsexColor
ersexColors   <- params$ersexColors

species  <- if (!is.null(params$taxon) && !is.na(params$taxon)) params$taxon else "Unknown taxon"
method   <- if (identical(params$method, "hull")) "Convex hull" else "Buffer"
buffer_km <- params$bufferDist

# Rows analysed spatially (have coordinates) and the full working dataset
points  <- params$points
col_name <- "Current Germplasm Type"
points[[col_name]] <- tidyr::replace_na(points[[col_name]], "H")
records <- if (is.data.frame(params$records) && nrow(params$records) > 0) params$records else points
if (!"source" %in% names(records)) records$source <- "unknown"
if (!"source" %in% names(points))  points$source  <- "unknown"
records[[col_name]] <- tidyr::replace_na(records[[col_name]], "H")

g_points <- points %>% filter(.data[[col_name]] == "G")
h_points <- points %>% filter(.data[[col_name]] == "H")

# Scores (FCS and priority were computed once by the app and passed in)
srs <- if (!is.null(params$srsMetrics) && "SRS exsitu" %in% names(params$srsMetrics)) params$srsMetrics$`SRS exsitu` else 0
grs <- if (!is.null(params$grsMetrics) && "GRS exsitu" %in% names(params$grsMetrics)) params$grsMetrics$`GRS exsitu` else 0
ers <- if (!is.null(params$ersMetrics$summary) && "ERS exsitu" %in% names(params$ersMetrics$summary)) params$ersMetrics$summary$`ERS exsitu` else 0
fcs <- if (!is.null(params$fcs) && !is.na(params$fcs)) params$fcs else mean(c(srs, grs, ers), na.rm = TRUE)
rank <- params$priority
rank_color <- params$priorityColor

fmt <- function(x, d = 1) formatC(round(x, d), format = "f", digits = d, big.mark = ",")
fmt0 <- function(x) formatC(x, format = "d", big.mark = ",")
```

```{r summary-cards, results='asis'}
card <- function(label, value, sub, color) {
  sprintf('<div class="score-card" style="background:%s"><div class="label">%s</div><div class="value">%s</div><div class="sub">%s</div></div>',
          color, label, value, sub)
}
cat('<div class="score-cards">',
    card("SRS ex situ", fmt(srs), "Sampling representativeness", combinedColor[1]),
    card("GRS ex situ", fmt(grs), sprintf("Geographic representativeness (%s km buffer)", buffer_km), grsexColor),
    card("ERS ex situ", fmt(ers), "Ecological representativeness", ersexColors[2]),
    card("FCS ex situ", fmt(fcs), "Final conservation score", "#2c3e50"),
    sprintf('<div class="score-card priority-card" style="background:%s"><div class="label">Priority category</div><div class="value" style="font-size:1.35rem">%s</div><div class="sub">based on FCS ex situ</div></div>', rank_color, rank),
    '</div>', sep = "\n")

n_gbif <- sum(records$source == "GBIF"); n_upload <- sum(records$source == "upload")
cat('<div class="facts">',
    sprintf('<div><span>Taxon</span><em>%s</em></div>', htmltools::htmlEscape(species)),
    sprintf('<div><span>Range estimate</span>%s</div>', if (method == "Buffer") sprintf("Buffer method (%s km around every record)", buffer_km) else sprintf("Convex hull of all records; %s km buffer around G records", buffer_km)),
    sprintf('<div><span>Records analysed</span>%s total, %s with coordinates</div>', fmt0(nrow(records)), fmt0(nrow(points))),
    sprintf('<div><span>Data sources</span>%s GBIF, %s uploaded</div>', fmt0(n_gbif), fmt0(n_upload)),
    sprintf('<div><span>Germplasm / reference</span>%s G, %s H</div>', fmt0(sum(records[[col_name]] == "G")), fmt0(sum(records[[col_name]] == "H"))),
    '</div>', sep = "\n")
```

# Occurrence records

The table below summarises the occurrence records used for the taxon, overall and by data source. A record "with coordinates" has a complete latitude and longitude pair and was used in the spatial analyses (GRS and ERS); the sampling score (SRS) uses every record. Each record was classified as an existing *ex situ* germplasm sample from a genebank, botanic garden, or other repository (**G**), or as a reference observation / voucher specimen (**H**). Records pulled from GBIF are labelled G when their basis of record is *living specimen*; uploaded records use the type supplied in the file.

```{r occurrence-table}
has_coords <- !is.na(records$Latitude) & !is.na(records$Longitude)
summarise_rows <- function(df, hc, label) {
  tibble(
    Source = label,
    `Total records` = nrow(df),
    `With coordinates` = sum(hc),
    `G records` = sum(df[[col_name]] == "G"),
    `G with coordinates` = sum(df[[col_name]] == "G" & hc),
    `H records` = sum(df[[col_name]] == "H"),
    `H with coordinates` = sum(df[[col_name]] == "H" & hc)
  )
}
occ <- bind_rows(
  summarise_rows(records, has_coords, "All records"),
  if (n_gbif > 0) summarise_rows(records[records$source == "GBIF", ], has_coords[records$source == "GBIF"], "GBIF"),
  if (n_upload > 0) summarise_rows(records[records$source == "upload", ], has_coords[records$source == "upload"], "Uploaded")
)
kbl(occ, format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  row_spec(1, bold = TRUE)
```

# Conservation scores

The table and chart show the three *ex situ* conservation gap analysis scores and their average, the Final Conservation Score (FCS *ex situ*). All scores range from 0 (poor state of conservation) to 100 (comprehensive conservation). Based on an FCS *ex situ* of **`r fmt(fcs)`**, the taxon is assigned the priority category **`r rank`**.

```{r scores-table}
scores_tbl <- tibble(
  Taxon = species,
  `Total records` = nrow(records),
  `Records with coordinates` = nrow(points),
  `SRS ex situ` = round(srs, 1),
  `GRS ex situ` = round(grs, 1),
  `ERS ex situ` = round(ers, 1),
  `FCS ex situ` = round(fcs, 1),
  `Priority category` = rank
)
kbl(scores_tbl, format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(8, background = rank_color)
```

```{r scores-plot, fig.align='center', out.width="100%"}
df_plot <- tibble(
  Metric = factor(c("SRS ex situ", "GRS ex situ", "ERS ex situ", "FCS ex situ"),
                  levels = c("SRS ex situ", "GRS ex situ", "ERS ex situ", "FCS ex situ")),
  Score = round(c(srs, grs, ers, fcs), 1),
  Color = c(combinedColor[1], grsexColor, ersexColors[2], "#2c3e50")
)
plot_ly(df_plot, x = ~Metric, y = ~Score, type = "bar", marker = list(color = ~Color),
        text = ~Score, textposition = "outside", hovertemplate = "%{x}: %{y}<extra></extra>",
        height = 380) %>%
  layout(
    yaxis = list(title = "Score (0-100)", range = c(0, 108), tickvals = c(0, 25, 50, 75, 100)),
    xaxis = list(title = ""),
    shapes = lapply(c(25, 50, 75), function(y) list(type = "line", x0 = -0.5, x1 = 3.5, y0 = y, y1 = y,
                                                   line = list(color = "#adb5bd", width = 1, dash = "dot"))),
    annotations = list(
      list(x = 3.5, y = 12.5, text = "UP", showarrow = FALSE, xanchor = "left", font = list(color = "#6c757d", size = 11)),
      list(x = 3.5, y = 37.5, text = "HP", showarrow = FALSE, xanchor = "left", font = list(color = "#6c757d", size = 11)),
      list(x = 3.5, y = 62.5, text = "MP", showarrow = FALSE, xanchor = "left", font = list(color = "#6c757d", size = 11)),
      list(x = 3.5, y = 87.5, text = "LP", showarrow = FALSE, xanchor = "left", font = list(color = "#6c757d", size = 11))
    ),
    margin = list(t = 20, r = 40)
  ) %>%
  config(displayModeBar = FALSE)
```

# Map of *ex situ* conservation gap analysis measures

`r if (method == "Convex hull") paste0("The wild range was estimated as the **convex hull** of all records with coordinates, clipped to land, and is drawn as a dashed outline. Germplasm (G) records were buffered by ", buffer_km, " km within that range to estimate the area already collected.") else paste0("The wild range was estimated with the **buffer method**: a ", buffer_km, " km buffer around every record with coordinates, clipped to land. The buffers around germplasm (G) records estimate the area already collected.")`

The **GRS gap** layer shows the part of the estimated range that lies farther than `r buffer_km` km from any germplasm record. The **ERS regions** layer shows the ecoregions occupied by the range, coloured by whether they overlap a germplasm buffer (covered) or not (gap). Records can be clicked for their attributes; layers can be toggled with the control in the top right, and the global protected-area layer (World Database on Protected Areas) can be switched on to see where collecting permits may be required.

```{r map}
m <- leaflet(width = "100%", height = 560) %>%
  # Esri tiles only: OpenStreetMap's tile servers refuse requests from a saved
  # HTML file opened locally (HTTP 403), which is how this report is used.
  addProviderTiles("Esri.WorldGrayCanvas", group = "Light Gray") %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topography") %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery") %>%
  addTiles(
    urlTemplate = "https://data-gis.unep-wcmc.org/server/rest/services/ProtectedSites/The_World_Database_of_Protected_Areas/MapServer/tile/{z}/{y}/{x}",
    group = "Protected areas (WDPA)",
    attribution = "Protected areas: UNEP-WCMC and IUCN, Protected Planet (WDPA)",
    options = tileOptions(opacity = 0.6, maxNativeZoom = 15, maxZoom = 19)
  ) %>%
  addMapPane("ersGap", zIndex = 400) %>%
  addMapPane("grsGap", zIndex = 405) %>%
  addMapPane("buffers", zIndex = 410) %>%
  addMapPane("points", zIndex = 420)

if (!is.null(params$sf_model) && inherits(params$sf_model, "sf") && nrow(params$sf_model) > 0) {
  model_simplified <- rmapshaper::ms_simplify(params$sf_model, keep = 0.3, keep_shapes = TRUE)
  m <- m %>% addPolygons(
    data = model_simplified, group = "Range (convex hull)",
    color = combinedColor[1], fillColor = combinedColor[1], fillOpacity = 0.15, weight = 2, dashArray = "6 4",
    options = pathOptions(pane = "buffers"), popup = "Estimated range (convex hull of all records)"
  )
}

if (!is.null(params$sf_buffers) && inherits(params$sf_buffers, "sf") && nrow(params$sf_buffers) > 0) {
  buffers_simplified <- rmapshaper::ms_simplify(params$sf_buffers, keep = 0.3, keep_shapes = TRUE)
  pal_type <- colorFactor(c("H" = combinedColor[1], "G" = combinedColor[2]), domain = c("H", "G"))
  m <- m %>% addPolygons(
    data = buffers_simplified, group = "Buffers",
    color = pal_type(buffers_simplified$processing_type),
    fillColor = pal_type(buffers_simplified$processing_type),
    fillOpacity = 0.4, weight = 1, options = pathOptions(pane = "buffers"),
    popup = paste0("<b>", ifelse(buffers_simplified$processing_type == "G", "Germplasm", "Reference"), " buffer</b><br/>", buffer_km, " km")
  )
}

if (!is.null(params$sf_grs_gap) && inherits(params$sf_grs_gap, "sf")) {
  grs_gap_simplified <- rmapshaper::ms_simplify(params$sf_grs_gap, keep = 0.3, keep_shapes = TRUE)
  m <- m %>% addPolygons(
    data = grs_gap_simplified, group = "GRS gap",
    color = grsexColor, fillColor = grsexColor,
    fillOpacity = 0.4, weight = 1, options = pathOptions(pane = "grsGap"),
    popup = "Geographic gap: range farther than the buffer distance from any germplasm record"
  )
}

if (!is.null(params$sf_ers_regions) && inherits(params$sf_ers_regions, "sf")) {
  ers_regions_simplified <- rmapshaper::ms_simplify(params$sf_ers_regions, keep = 0.3, keep_shapes = TRUE)
  pal_ers <- colorFactor(c("Covered" = ersexColors[1], "Gap Ecoregion" = ersexColors[2]), domain = c("Covered", "Gap Ecoregion"))
  m <- m %>% addPolygons(
    data = ers_regions_simplified, group = "ERS regions",
    color = "black", fillColor = ~pal_ers(gap_status),
    fillOpacity = 0.3, weight = 1.5, options = pathOptions(pane = "ersGap"),
    popup = ~paste0("<b>", ECO_NAME, "</b><br/>", ifelse(gap_status == "Covered", "Covered by a germplasm buffer", "Ecological gap"))
  )
}

gen_popup <- function(data) {
  paste0(
    "<strong>", htmlEscape(as.character(data$`Taxon Name`)), "</strong>",
    "<br/><b>Type:</b> ", data[[col_name]], " &nbsp; <b>Source:</b> ", data$source,
    if ("Accession Number" %in% names(data)) paste0("<br/><b>Accession:</b> ", htmlEscape(as.character(data$`Accession Number`))) else "",
    if ("Collection Date" %in% names(data)) paste0("<br/><b>Date:</b> ", htmlEscape(as.character(data$`Collection Date`))) else "",
    "<br/><b>Collector:</b> ", htmlEscape(as.character(data$Collector)),
    "<br/><b>Locality:</b> ", htmlEscape(as.character(data$Locality))
  )
}

if (nrow(h_points) > 0) {
  m <- m %>% addCircleMarkers(
    data = h_points, lng = ~Longitude, lat = ~Latitude, group = "Reference records (H)",
    radius = 4, color = "#4d4d4d", fillColor = combinedColor[1], fillOpacity = 0.85, weight = 1,
    popup = gen_popup(h_points), popupOptions = popupOptions(maxWidth = 300), options = pathOptions(pane = "points")
  )
}
if (nrow(g_points) > 0) {
  m <- m %>% addCircleMarkers(
    data = g_points, lng = ~Longitude, lat = ~Latitude, group = "Germplasm records (G)",
    radius = 4, color = "#4d4d4d", fillColor = combinedColor[2], fillOpacity = 0.85, weight = 1,
    popup = gen_popup(g_points), popupOptions = popupOptions(maxWidth = 300), options = pathOptions(pane = "points")
  )
}

if (nrow(points) > 0) {
  m <- m %>% fitBounds(min(points$Longitude), min(points$Latitude), max(points$Longitude), max(points$Latitude))
}

overlays <- c("Reference records (H)", "Germplasm records (G)",
              if (method == "Convex hull") "Range (convex hull)", "Buffers", "GRS gap", "ERS regions",
              "Protected areas (WDPA)")
m %>%
  addLayersControl(baseGroups = c("Light Gray", "Topography", "Imagery"), overlayGroups = overlays,
                   options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup(c("ERS regions", "Protected areas (WDPA)")) %>%
  addLegend(position = "bottomleft", colors = c(combinedColor[1], combinedColor[2], grsexColor, ersexColors[1], ersexColors[2]),
            labels = c("Reference records / buffers (H)", "Germplasm records / buffers (G)", "GRS gap", "ERS: covered ecoregion", "ERS: gap ecoregion"),
            opacity = 0.8, title = "Legend")
```

## *Ex situ* conservation ecoregions table

All terrestrial ecoregions occupied by the estimated range of the taxon. **Gap ecoregion** is *true* when the ecoregion overlaps the range but lies farther than `r buffer_km` km from any germplasm record, i.e. it is not represented in *ex situ* collections. The record counts are the analysed records (with coordinates) falling inside each ecoregion.

```{r ecoregions-table}
if (!is.null(params$sf_ers_regions) && inherits(params$sf_ers_regions, "sf") && nrow(params$sf_ers_regions) > 0) {
  eco <- params$sf_ers_regions
  pts_sf <- sf::st_as_sf(points, coords = c("Longitude", "Latitude"), crs = sf::st_crs(eco), remove = FALSE)
  pts_sf$.id <- seq_len(nrow(pts_sf))
  joined <- suppressWarnings(sf::st_join(pts_sf, eco[, "ECO_NAME"], join = sf::st_intersects)) %>%
    sf::st_drop_geometry() %>%
    filter(!is.na(ECO_NAME)) %>%
    distinct(.id, ECO_NAME, .keep_all = TRUE) %>%
    group_by(ECO_NAME) %>%
    summarise(`G records` = sum(.data[[col_name]] == "G"), `H records` = sum(.data[[col_name]] == "H"), .groups = "drop")

  eco_table <- eco %>%
    sf::st_drop_geometry() %>%
    distinct(ECO_NAME, gap_status) %>%
    mutate(`Gap ecoregion` = gap_status != "Covered") %>%
    left_join(joined, by = "ECO_NAME") %>%
    mutate(across(c(`G records`, `H records`), ~ tidyr::replace_na(.x, 0L))) %>%
    select(`Gap ecoregion`, `Ecoregion name` = ECO_NAME, `G records`, `H records`) %>%
    arrange(desc(`Gap ecoregion`), `Ecoregion name`)

  DT::datatable(eco_table, rownames = FALSE, class = "compact stripe hover",
                options = list(pageLength = 10, order = list())) %>%
    DT::formatStyle("Gap ecoregion", backgroundColor = DT::styleEqual(c(TRUE, FALSE), c(paste0(ersexColors[2], "66"), paste0(ersexColors[1], "66"))))
} else {
  cat("No ecoregion data available for this analysis.")
}
```

## Records used in the analysis

Every record with coordinates that entered the spatial analysis. GBIF accession numbers link to the occurrence page on gbif.org.

```{r records-table}
rec <- points %>%
  mutate(
    `Accession Number` = ifelse(source == "GBIF" & !is.na(`Accession Number`),
      sprintf('<a href="https://www.gbif.org/occurrence/%s" target="_blank">%s</a>', `Accession Number`, `Accession Number`),
      as.character(`Accession Number`)),
    Latitude = round(Latitude, 4), Longitude = round(Longitude, 4)
  ) %>%
  select(any_of(c("Accession Number", "Taxon Name", "Current Germplasm Type", "source", "Collection Date",
                  "Latitude", "Longitude", "Locality", "Collector"))) %>%
  rename(Type = `Current Germplasm Type`, Source = source)
DT::datatable(rec, rownames = FALSE, escape = FALSE, class = "compact stripe hover",
              filter = "top", options = list(pageLength = 10, scrollX = TRUE))
```

# Definitions of conservation gap analysis scores

### Sampling Representativeness Score (SRS *ex situ*)

The ratio of germplasm accessions (G) available in *ex situ* repositories to reference / voucher (H) records for the taxon, using all compiled records regardless of whether they include coordinates. If there are more G than H records the score is capped at 100.

### Geographic Representativeness Score (GRS *ex situ*)

Buffers of `r buffer_km` km are created around each G record to estimate the geographic area already collected within the estimated range of the taxon; the score is the proportion of the range covered by these buffers. The range is estimated either as the union of `r buffer_km` km buffers around all records (buffer method) or as the convex hull of all records clipped to land (convex hull method).

### Ecological Representativeness Score (ERS *ex situ*)

The proportion of terrestrial ecoregions represented within the G buffers out of the total number of ecoregions occupied by the estimated range. Ecoregions follow Olson et al. (2001), 814 terrestrial ecoregions worldwide.

### Final Conservation Score (FCS *ex situ*)

The average of the three *ex situ* scores. Taxa are categorised as **Urgent Priority (UP)** for further conservation action when FCS < 25, **High Priority (HP)** when 25 ≤ FCS < 50, **Medium Priority (MP)** when 50 ≤ FCS < 75, and **Low Priority (LP)** when FCS ≥ 75.

# Definitions of occurrence data categories

**Total records**: the number of observations and *ex situ* collection records for the taxon evaluated in this analysis.

**Records with coordinates**: records with a valid latitude and longitude pair, used in the spatial analyses.

**Germplasm records (G)**: occurrences for which a living sample (typically a plant or seed) is held in an *ex situ* conservation system (botanic garden, seed bank, genebank, etc.).

**Reference / voucher records (H)**: occurrences supported by an herbarium specimen or other reference record.

```{=html}
<div class="footer-note">
<strong>Methodology.</strong> Metrics follow the <a href="https://github.com/CIAT-DAPA/GapAnalysis" target="_blank">GapAnalysis R package</a> (Carver et al. 2021, <em>Ecography</em>) and Khoury et al. (2019, <em>Ecological Indicators</em>). Occurrence data from GBIF are subject to the <a href="https://www.gbif.org/terms" target="_blank">GBIF data user agreement</a>; protected-area tiles from UNEP-WCMC and IUCN, <a href="https://www.protectedplanet.net" target="_blank">Protected Planet</a>. This report is intended for conservation planning and should be validated with local experts and field assessments.
</div>
```
