---
title: "Ex situ Gap Analysis Report"
output: html_document
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  points: NA
  bufferDist: NA
  srsMetrics: NA
  grsMetrics: NA
  ersMetrics: NA
  sf_buffers: NA
  sf_grs_gap: NA
  sf_ers_regions: NA
  combinedColor: NA
  grsexColor: NA
  ersexColors: NA
---

```{=html}
<style>
  @import url('https://fonts.googleapis.com/css2?family=Roboto:wght@400;500;700&display=swap');

  div.main-container {
    max-width: 1600px !important;
  }

  .footer {
    position: fixed;
    left: 0;
    bottom: 0;
    width: 100%;
    background-color: red;
    color: white;
    text-align: center;
  }

  body {
    font-family: 'Roboto', sans-serif;
    padding: 0 40px; /* 0 for top and bottom, 40px for left and right */
  }

  h1 {
    font-size: 24px;
    font-weight: 500;
    margin-bottom: 16px;
    background-color: #f0f0f0;
    border-radius: 30px;
    padding: 10px;
    text-align: center;
  }

  h2 {
    font-size: 20px;
    font-weight: 500;
    margin-bottom: 12px;
    background-color:#f0f0f0;
    border-radius: 15px;
    padding: 8px;
  }

  h3 {
    font-size: 16px;
    font-weight: 500;
    margin-bottom: 8px;
    background-color: #f0f0f0;
    padding: 8px;
  }

  p {
    font-size: 14px;
    font-weight: 400;
    line-height: 1.5;
    margin: 0 auto;
    width: 90%;
    margin-bottom: 8px;
  }

  a {
    color: #007bff;
    text-decoration: none;
  }

  div.datatables {
    color: #333;
    background: #eceae640;
  }
</style>
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
pacman::p_load(leaflet, dplyr, ggplot2, DT, sf, htmltools, rmapshaper)

# Force DT dependencies to load
htmltools::tagList(DT::datatable(cars[1,1,drop=FALSE], width=0, height=0, options=list(dom='t', autoWidth=TRUE)))

# Extract Colors from params
combinedColor <- params$combinedColor
grsexColor <- params$grsexColor
ersexColors <- params$ersexColors

# Extract Parameters
points <- params$points
col_name <- if ("Current Germplasm Type" %in% names(points)) "Current Germplasm Type" else "type"
points[[col_name]] <- tidyr::replace_na(points[[col_name]], "H")

species <- if(nrow(points) > 0 && !is.null(points$`Taxon Name`)) points$`Taxon Name`[1] else "Unknown Species"
totalPoints <- nrow(points)
g_points <- points %>% filter(.data[[col_name]] == "G")
h_points <- points %>% filter(.data[[col_name]] == "H")

# Unpack Scores
srs <- if(!is.null(params$srsMetrics) && "SRS exsitu" %in% names(params$srsMetrics)) params$srsMetrics$`SRS exsitu` else 0
grs <- if(!is.null(params$grsMetrics) && "GRS exsitu" %in% names(params$grsMetrics)) params$grsMetrics$`GRS exsitu` else 0
ers <- if(!is.null(params$ersMetrics$summary) && "ERS exsitu" %in% names(params$ersMetrics$summary)) params$ersMetrics$summary$`ERS exsitu` else 0
fcs <- mean(c(srs, grs, ers), na.rm = TRUE)

# Prioritization Rank
rank <- case_when(
  fcs <= 25 ~ "Urgent Priority (UP)",
  fcs > 25 & fcs <= 50 ~ "High Priority (HP)",
  fcs > 50 & fcs <= 75 ~ "Medium Priority (MP)",
  TRUE ~ "Low Priority (LP)"
)

rank_color <- case_when(
  rank == "Urgent Priority (UP)" ~ "#ffb4b3",
  rank == "High Priority (HP)" ~ "#ffd380",
  rank == "Medium Priority (MP)" ~ "#ffff80",
  TRUE ~ "#a8d2a8"
)
```

# `r paste0('*', species, '*')`

<br>

## Occurrence records    

The table below shows a summary of the occurrence records for the taxon. In preparation for the conservation gap analyses, we classified each record based on whether it was an existing *ex situ* germplasm sample from a genebank, botanic garden, or other repository (labeled G, as most records were from genebanks), or a reference observation/voucher specimen (labeled H, as most of these records were from herbaria).

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary_table <- data.frame(
  Taxon = species,
  `Total Occurrences` = totalPoints,
  `Germplasm Records (G)` = nrow(g_points),
  `Reference Records (H)` = nrow(h_points),
  check.names = FALSE
)

DT::datatable(summary_table, rownames = FALSE, class = "compact", options = list(dom = 't'))

```

<br>

## Map of *ex situ* conservation gap analysis measures

The GRS *ex situ* map feature shows the potential distribution model, with previous *ex situ* germplasm collection points surrounded by a buffer overlaid. Only germplasm (G) points and reference observations (H) are displayed on the map.

The ERS *ex situ* map feature shows ecoregions within the potential distribution model from which no *ex situ* germplasm collections have been made.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Base Map
m <- leaflet(width = "100%") %>%
  addProviderTiles("OpenStreetMap", group = "OpenStreetMap") %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topography") %>%
  addProviderTiles("Esri.WorldImagery", group = "Imagery") %>%
  addLayersControl(
    baseGroups = c("OpenStreetMap", "Topography", "Imagery"),
    overlayGroups = c("Reference Records", "Germplasm Records", "Buffers", "GRS Gap", "ERS Regions"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMapPane("ersGap", zIndex = 400) %>%
  addMapPane("grsGap", zIndex = 405) %>%
  addMapPane("buffers", zIndex = 410)

# Add pre-calculated sf objects (simplified for memory reduction)
if (!is.null(params$sf_buffers) && nrow(params$sf_buffers) > 0) {
  # Simplify geometries to save memory in the HTML output using rmapshaper
  buffers_simplified <- rmapshaper::ms_simplify(params$sf_buffers, keep = 0.05, keep_shapes = TRUE)
  pal_type <- colorFactor(c("H" = combinedColor[2], "G" = combinedColor[1]), domain = c("H", "G"))
  m <- m %>% addPolygons(
    data = buffers_simplified, group = "Buffers",
    color = pal_type(buffers_simplified$processing_type),
    fillColor = pal_type(buffers_simplified$processing_type),
    fillOpacity = 0.4, weight = 1, options = pathOptions(pane = "buffers"),
    popup = paste("Buffer:", params$bufferDist, "km")
  )
}

if (!is.null(params$sf_grs_gap)) {
  grs_gap_simplified <- rmapshaper::ms_simplify(params$sf_grs_gap, keep = 0.05, keep_shapes = TRUE)
  m <- m %>% addPolygons(
    data = grs_gap_simplified, group = "GRS Gap",
    color = grsexColor, fillColor = grsexColor,
    fillOpacity = 0.4, weight = 1, options = pathOptions(pane = "grsGap"),
    popup = "Geographic Gap"
  )
}

if (!is.null(params$sf_ers_regions)) {
  ers_regions_simplified <- rmapshaper::ms_simplify(params$sf_ers_regions, keep = 0.05, keep_shapes = TRUE)
  pal_ers <- colorFactor(c(ersexColors[1], ersexColors[2]), domain = ers_regions_simplified$gap_status)
  m <- m %>% addPolygons(
    data = ers_regions_simplified, group = "ERS Regions",
    color = "black", fillColor = ~pal_ers(gap_status),
    fillOpacity = 0.3, weight = 2, options = pathOptions(pane = "ersGap"),
    popup = ~paste("<b>Region:</b>", ECO_NAME, "<br><b>Status:</b>", gap_status)
  )
}

# Popup Generator
gen_popup <- function(data) {
  paste0(
    "<strong>", as.character(data$`Taxon Name`), "</strong>",
    "<br/><strong> Type: </strong>", data[[col_name]],
    "<br/><b>Collector Name:</b> ", data$Collector,
    "<br/><b>Locality Description:</b> ", data$Locality
  )
}

# Add Points
if (nrow(h_points) > 0) {
  m <- m %>% addCircleMarkers(
    data = h_points, lng = ~Longitude, lat = ~Latitude, group = "Reference Records",
    radius = 4, color = "white", fillColor = combinedColor[1], fillOpacity = 0.8, weight = 1,
    popup = ~gen_popup(h_points)
  )
}

if (nrow(g_points) > 0) {
  m <- m %>% addCircleMarkers(
    data = g_points, lng = ~Longitude, lat = ~Latitude, group = "Germplasm Records",
    radius = 4, color = "white", fillColor = combinedColor[2], fillOpacity = 0.8, weight = 1,
    popup = ~gen_popup(g_points)
  )
}

m
```

<br>

## Conservation scores 

The figure below displays all conservation gap analysis scores, providing a summary of the *ex situ* conservation scores. Based on the Final Conservation Score (FCS), the taxon has an assigned priority category of: **`r rank`**.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="100%"}
df_plot <- tibble(
  Metric = factor(c("SRS", "GRS", "ERS", "FCS"), levels = c("SRS", "GRS", "ERS", "FCS")),
  Score = c(srs, grs, ers, fcs),
  Type = c("SRS", "GRS", "ERS", "FCS")
) %>% mutate(Score = round(Score, 1))

df_plot$x_label <- factor(paste0(df_plot$Metric, "\n", df_plot$Score), levels = paste0(df_plot$Metric, "\n", df_plot$Score))

ggplot(df_plot, aes(x = x_label, y = Score, fill = Type)) +
  geom_col(width = 0.6) +
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100), expand = c(0, 0)) +
  scale_fill_manual(values = c("SRS" = combinedColor[1], "GRS" = grsexColor, "ERS" = ersexColors[2], "FCS" = "#2c3e50")) +
  theme_minimal(base_size = 14) +
  labs(y = "Score (0-100)", x = "") +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold", hjust = 0.5)
  )

```

<br>

### *Ex situ* conservation ecoregions table

The table below shows all the ecoregions overlapping with the reference occurrence area of the taxon. The column ERS *ex situ* gap ecoregion denotes (true) if an ecoregion overlaps with the expected area but is farther than the buffer distance from any *ex situ* conservation occurrence, and is thus considered a gap in *ex situ* conservation.

```{r echo=FALSE, message=FALSE, warning=FALSE}
if (!is.null(params$sf_ers_regions) && nrow(params$sf_ers_regions) > 0) {
  eco_table <- params$sf_ers_regions %>%
    st_drop_geometry() %>%
    dplyr::mutate(
      `ERS ex situ gap ecoregion` = gap_status != "Covered"
    ) %>%
    dplyr::select(
      `ERS ex situ gap ecoregion`,
      `Ecoregion name` = ECO_NAME
    ) %>%
    dplyr::arrange(desc(`ERS ex situ gap ecoregion`))
  
  DT::datatable(eco_table, rownames = FALSE, class = "compact", options = list(pageLength = 10))
} else {
  cat("No ecoregion data available for this analysis.")
}
```

<br>



<br>

## Definitions of conservation gap analysis scores

### Sampling Representativeness Score (SRS)

***Ex situ***: The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher/herbarium (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

### Geographic Representativeness Score (GRS)

***Ex situ***: The Geographic Representativeness Score *ex situ* (GRS ex situ) uses buffers created around each G collection coordinate point to estimate geographic areas already well collected within the potential distribution models of each taxon and then calculates the proportion of the potential distribution model covered by these buffers.

### Ecological Representativeness Score (ERS)

***Ex situ***: The Ecological Representativeness Score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the potential distribution model.

### Final Conservation Score (FCS)

***Ex situ***: The Final Conservation Score *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics.

Taxa were further categorized with regard to conservation priority, with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥ 75.

## Definitions of occurrence data categories

**Occurrences**: The total number of observations and *ex situ* collections records for the taxon that were evaluated for use within the study.

**Occurrences with lat/long**: The total number of observations that had a valid latitude and longitude pair and were used for spatial analysis.

**Germplasm Records (G)**: Occurrences in which a living sample (typically a live plant or seed) is present in an *ex situ* conservation system (i.e., botanical garden, seed bank, genebank, etc.).

**Reference/voucher Records (H)**: Occurrences that have a supporting herbarium or other reference record.

---

<p style="font-size: 12px; text-align: center; color: #777;"><strong>Methodology Note:</strong> These metrics are derived from the <a href="https://github.com/CIAT-DAPA/GapAnalysis">GapAnalysis R Package</a> guidelines. This report is intended for conservation planning purposes and should be validated with local experts and field assessments.</p>